Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-10-30 main.tex
Started updating text in seciond 2.2 for the new Fig 2.

\section*{Highlights}
%
\begin{itemize}
  \item A bicycle steering motor with automatic roll rate feedback control can
    stabilize the bicycle-rider system at speeds as low as 4.6~\si{\kph}.
  \item Every bicycle rider has their own perturbation resistance threshold
    when subjected to mechanical handlebar perturbations.
  \item Automatic balance assistance reduces the probability of falling while
    riding at 6~\si{\kph} when mechanically perturbed.
  \item Balance assist technology can reduce natural single-actor falls around
    the limits of the rider's control authority.
\end{itemize}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a Teensy microprocessor (PJRC, Sherwood, USA) and data from all
sensors is logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhøj, Denmark)
at least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use the gain factor \(\kappa=8\) (low) and \(\kappa=10\) (high) to tune the gain
magnitude during the experiments. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled with \(\kappa=8\) (lower
    row) root locus of the eigenvalue components (real: solid, imaginary:
    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
    without the inertial effect of a rigid rider (left) and with a rigid rider
    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
    region is the linear stable speed range. Geometric and inertial parameters
    for these plots were estimated with the methods presented in
    \citet{Moore2012} and software packages
    BicycleParameters~1.1.1~\citep{Moore2024} and
    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
    Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

>>>>

\section*{Highlights}
%
\begin{itemize}
  \item A bicycle steering motor with automatic roll rate feedback control can
    stabilize the bicycle-rider system at speeds as low as 8~\si{\kph}.
  \item Every bicycle rider has their own perturbation resistance threshold
    when subjected to mechanical handlebar perturbations.
  \item Automatic balance assistance reduces the probability of falling while
    riding at 6~\si{\kph} when mechanically perturbed.
  \item Balance assist technology can reduce natural single-actor falls around
    the limits of the rider's control authority.
\end{itemize}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a Teensy microprocessor (PJRC, Sherwood, USA) and data from all
sensors is logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhøj, Denmark)
at at least 100 Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune
the gain magnitude during the experiments. For the 6~\si{\kilo\meter\per\hour}
experiment, scaling the proportional feedback gain linearly with respect to
speed stabilizes the normally unstable weave mode of the bicycle down to
3.4~\si{\kilo\meter\per\hour} for the riderless bicycle and
9.7~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Geometric and inertial parameters for these plots
were estimated with the methods presented in \citet{Moore2012} and software
packages BicycleParameters~1.1.1~\citep{Moore2024} and
Yeadon~1.5.0~\citep{Dembia2015} and are shown in
Appendix~\ref{sec:physical-parameters}. The identified weave mode uses the
methods of Kooijman et. al~\citep{Kooijman2009a} and gives an indication of the
performance of the system.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controleld with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
  frequency.}
  \label{fig:eigenvalues}
\end{figure}
